---
title: |
  | Study - UTR Widths
subtitle: |
  | Endome Generation Pipeline
author:
- name: "Guillermo Rocamora Pérez"
  affiliation: UCL - Version 1.0
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  bookdown::html_document2:
    figure_caption: yes
    code_folding: hide
    theme: paper
    highlight: pygments
    # highlight: haddock
    df_print: paged
    toc: true
    toc depth: 3
    toc_float: true
    number_sections: true
  md_document:
    variant: markdown_github
    toc: true
    number_sections: true
always_allow_html: true
link-citations: true
notes-after-punctuation: false
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/dependent/jhep-reports.csl
editor_options: 
  chunk_output_type: inline
---

```{R setup, echo = F, message = F, warning = F}
# Load libraries
library(rtracklayer)
library(GenomicFeatures)
library(tidyverse)

# Define main path
main_path <- file.path("/home/grocamora/RytenLab-Research/38-Endome_generation")# here::here()

# Source helper functions
source(file.path(main_path, "R/hf_graph_and_themes.R"))
source(file.path(main_path, "R/hf_notebooks.R"))

# Set options
options(dplyr.summarise.inform = FALSE)
options(lifecycle_verbosity = "warning")
knitr::opts_chunk$set(echo = T, warning = F, message = F, out.width="100%", 
                      fig.align = "center", dpi = 300, 
                      fig.height = 7.2, fig.width = 7.2)
```

```{r setup-theme, echo = F, results = "asis"}
## Custom sciRmdTheme with increased width.
## Please ask guillermorocamora@gmail.com for more details.
## Download from devtools::install_github("https://github.com/guillermo1996/grpSciRmdTheme")
if(require(grpSciRmdTheme)){
  grpSciRmdTheme::set.theme(
    theme = "default",
    color = NULL,
    header.sticky = FALSE,
    list.group.icon = "arrow",
    font.family = "Arial",
    font.color = "black",
    header.color = "darkblue",
    figure.perc = 100
  )
}

## A bug in recent versions of KableExtra requires to load the HTML classes in
## some circumstances.
kableExtra::html_dependency_lightable()

## Custom styles for the output html
cat('
<style type="text/css">
</style>')
```

# Introduction

> Aim: extract the length of annotated UTRs in GENCODE v47

As part of the *Endome* generation pipeline, the transcriptome truncation requires as a user input parameter the truncation width - basically, the length of the start or end sequence to extract from each transcript.

According to the work of [Mervin M. Fansler (2024)](https://www.nature.com/articles/s41467-024-48254-9) with `scUTRquant`, the cut-off for truncation was empirically observed to be 500 nucleotides since more than 95% of 3' end reads of tested reference genes map within that limit. Their data was extracted from Tabula Muris scRNA-seq data of 28 libraries and 56 curated genes.

# Load & pre-process data

Import the reference annotation from GENCODE v47.

```{r, load-gencode}
gencode_path <- "/home/grocamora/RytenLab-Research/Additional_files/GENCODE/gencode.v47.annotation.gtf"

gencode_ref <- rtracklayer::import.gff(gencode_path)
txdb_ref <- GenomicFeatures::makeTxDbFromGFF(gencode_path)
```

Next, filter the imported reference annotation to contain only protein-coding transcripts:

```{r}
pc_gencode_ref <- gencode_ref %>% plyranges::filter(type == "transcript", transcript_type == "protein_coding")
```

Additionally, the transcripts from blacklisted regions are removed:

```{r, load-blacklisted}
blacklisted_regions <- "/home/grocamora/RytenLab-Research/Additional_files/hg38-blacklist.v2.bed"
blacklisted_regions <- rtracklayer::import.bed(blacklisted_regions)

blacklisted_overlaps <- GenomicRanges::findOverlaps(pc_gencode_ref, blacklisted_regions, type = "any")
pc_gencode_filter_ref <- tibble::as_tibble(pc_gencode_ref[-queryHits(blacklisted_overlaps)])
```

We will also remove transcripts with duplicated coordinates:

```{r}
clean_df <- pc_gencode_filter_ref %>% 
  dplyr::group_by(seqnames, start, end, strand, gene_id) %>% 
  dplyr::arrange(transcript_support_level, transcript_id, .by_group = T) %>% 
  dplyr::slice_head(n = 1) %>% 
  dplyr::ungroup() %>% 
  dplyr::mutate(transcript_support_level = as.numeric(na_if(transcript_support_level, "NA")))
```

# Extract 3' and 5' lengths

```{r}
x3_utr <- unlist(GenomicFeatures::threeUTRsByTranscript(txdb_ref, use.names = T))
x5_utr <- unlist(GenomicFeatures::fiveUTRsByTranscript(txdb_ref, use.names = T))

# Extract 3' Widths
mcols(x3_utr)["transcript_id"] <- names(x3_utr)
tx_3width <- tibble::as_tibble(x3_utr) %>% 
  dplyr::group_by(transcript_id) %>% 
  dplyr::summarise(utr3_width = sum(width))

# Extract 5' Widths
mcols(x5_utr)["transcript_id"] <- names(x5_utr)
tx_5width <- tibble::as_tibble(x5_utr) %>% 
  dplyr::group_by(transcript_id) %>% 
  dplyr::summarise(utr5_width = sum(width))
```

## Add information to transcripts

```{r}
utr_width <- clean_df %>% 
  dplyr::left_join(tx_5width, by = "transcript_id") %>% 
  dplyr::left_join(tx_3width, by = "transcript_id") %>% 
  dplyr::select(seqnames, start, end, width, strand, 
                utr5_width, utr3_width,
                tx_id = transcript_id, tx_type = transcript_type, TSL = transcript_support_level, 
                gene_id, gene_name, tag)
```

# Results of the data 

## Split by TSL

In this section, we focus on the distribution of UTR widths for the different TSL available in the data. Results will showcase from TSL 1 to 5, but only the first three levels will be considered in the summary and in different sections.

<p style="font-size:14pt; text-decoration:underline;">Summary of results:</p>

+ For the 3' UTR end we observe a decrease in the median length of the UTRs as the TSL increases. The interquartile range also decreases significantly, meaning that the distributions become more a more compressed around the median. 
+ For the 5' UTR end we do not observe the previous behaviour. Both the median and the IQR stay almost constant, regardless of the TSL. Median values are also considerable smaller than for the 3' end, as expected.
+ In terms of missing widths, as the TSL increases we do observe an increase in the proportion of NA's on both end, as expected.

```{r plot-tsl, fig.width = 9, fig.height=5}
test_TSL <- c(1, 2, 3, 4, 5)
valid_TSL <- c(1, 2, 3)

utr_width %>% 
  dplyr::filter(TSL %in% test_TSL) %>% 
  dplyr::mutate(TSL = as.factor(TSL)) %>% 
  tidyr::pivot_longer(cols = c(utr5_width, utr3_width), names_to = "utr", values_to = "utr_width") %>% 
  ggplot(aes(x = utr, y = utr_width, fill = TSL)) +
  geom_boxplot() +
  geom_hline(aes(yintercept = 500, color = "500 bp"), linetype = 5, linewidth = 0.75) +
  scale_color_manual(name = "Txcutr cut-off", values = "black", guide = guide_legend(order = 1)) +
  ggsci::scale_fill_jco(name = "TSL", guide = guide_legend(order = 2)) +
  scale_y_log10() +
  scale_x_discrete(labels = c("utr3_width" = "3'", "utr5_width" = "5'")) +
  ggtitle("UTR Widths", subtitle = "By TSL") +
  labs(x = "UTR", y = "Width [bp]") +
  custom_gg_theme +
  theme(
    legend.position = "right",
    legend.box.just = "left",
    legend.spacing.y = unit(2, "lines")
  )
```

```{r table-tsl}
utr5_tsl <- lapply(test_TSL, function(iTSL) unclass(summary(utr_width %>% filter(TSL == iTSL) %>% pull(utr5_width)))) %>% 
  dplyr::bind_rows() %>% dplyr::mutate(TSL = test_TSL)
utr3_tsl <- lapply(test_TSL, function(iTSL) unclass(summary(utr_width %>% filter(TSL == iTSL) %>% pull(utr3_width)))) %>% 
  dplyr::bind_rows() %>% dplyr::mutate(TSL = test_TSL)

by_tsl_table <- dplyr::bind_rows(
  utr5_tsl %>% dplyr::mutate(UTR = "5"),
  utr3_tsl %>% dplyr::mutate(UTR = "3")
) %>% 
  dplyr::mutate(total = rep(utr_width$TSL %>% table(), 2)) %>% 
  dplyr::mutate(UTR = factor(UTR, levels = c("5", "3"))) %>% 
  dplyr::mutate(IQR = `3rd Qu.` - `1st Qu.`) %>% 
  dplyr::select(UTR, TSL,
                Median, IQR, `1st Qu.`, `3rd Qu.`, `NA's`, total)
  
by_tsl_table %>% 
  printPrettyDf(style = "md") %>% 
  kableExtra::pack_rows(index = table(by_tsl_table$UTR))
```

## Split by Gencode Tag

In this section, the median UTR width per Gencode tag is presented. In Gencode's GTF files, additional information of the transcripts are reported in the `tag` column. For more information about the possible tag values, please refer to their [documentation](https://www.gencodegenes.org/pages/tags.html).

Based on the distribution of categories, I decided to focus on only a subset of them: *CCDS*, *basic*, *exp_conf*, *GENCODE_Primary*, *alternative_5_UTR*, *alternative_3_UTR*, *cds_start_NF* and *cds_end_NF*. These categories alone already account for more than 97% of all transcripts.

<details>
<summary>Expand to see the number of transcripts per Gencode Tag</summary>
<br>
```{r tag-table}
utr_width %>% 
  dplyr::filter(TSL %in% valid_TSL) %>% 
  dplyr::group_by(tag) %>% 
  dplyr::count() %>% 
  dplyr::arrange(-n) %>% 
  `colnames<-`(c("GENCODE Tag", "Number of transcripts")) %>% 
  printPrettyDf(style = "html")
```
</details>
<br>

<details>
<summary>Expand to see Gencode Tag explanation</summary>
+ **CCDS**: member of the consensus CDS gene set, confirming coding regions between ENSEMBL, UCSC, NCBI and HAVANA.
+ **Basic**: identifies a subset of representative transcripts for each gene; prioritises full-length protein coding transcripts over partial or non-protein coding transcripts within the same gene, and intends to highlight those transcripts that will be useful to the majority of users.
+ **Exp_conf**: transcript was tested and confirmed experimentally.
+ **GENCODE_Primary**: belongs to a minimal set that contains MANE Select, MANE Plus Clinical and Ensembl Canonical transcripts and transcripts containing any conserved exons and common alternative splicing events (including exons skips) that are absent from the MANE and Ensembl Canonical transcripts for protein-coding genes. Other biotypes will have the GENCODE_Primary flag added to the Ensembl Canonical transcript and, for lncRNA genes only, this will be the transcript with the longest genomic span.
+ **Alternative_3_UTR**: shares an identical CDS but has alternative 5' UTR with respect to a reference variant.
+ **Alternative_5_UTR**: shares an identical CDS but has alternative 3' UTR with respect to a reference variant.
+ **Cds_end_NF**: the coding region end could not be confirmed.
+ **Cds_start_NF**: the coding region start could not be confirmed.
</details>
<br>

<p style="font-size:14pt; text-decoration:underline;">Summary of results:</p>

+ Since around 65% of the transcript are from the *CCDS* category, we will mostly focus on their results. The median UTR width in for the 3' end is around 750 nt, considerably above the 500 nt threshold by `txcutr`. 
+ For other categories, UTR length is substantially reduced. Most of them have a 3rd quantile (75% of the data) below the 500 nt cut-off.

```{r plot-tag, fig.width=12, fig.height=7}
valid_tags <- c("CCDS", "basic", "exp_conf", "GENCODE_Primary", "alternative_5_UTR", "alternative_3_UTR", "cds_start_NF", "cds_end_NF")

utr_tag_width <- utr_width %>% 
  dplyr::filter(TSL %in% valid_TSL) %>% 
  dplyr::filter(tag %in% valid_tags) %>% 
  dplyr::mutate(TSL = as.factor(TSL)) %>% 
  dplyr::mutate(tag = factor(tag, levels = valid_tags))

tag_names <- utr_tag_width$tag %>% table %>% c()
tag_labels <- setNames(paste0(names(tag_names), " (n = ", tag_names, ")"), names(tag_names))

utr_tag_width %>% 
  tidyr::pivot_longer(cols = c(utr5_width, utr3_width), names_to = "utr", values_to = "utr_width") %>% 
  ggplot(aes(x = utr, y = utr_width, fill = tag)) +
  geom_boxplot() +
  geom_hline(aes(yintercept = 500, color = "500 bp"), linetype = 5, linewidth = 0.75) +
  scale_color_manual(name = "Txcutr cut-off", values = "black", guide = guide_legend(order = 1)) +
  ggsci::scale_fill_jco(name = "Gencode Tag", guide = guide_legend(order = 2)) +
  scale_y_log10() +
  scale_x_discrete(labels = c("utr3_width" = "3'", "utr5_width" = "5'")) +
  ggh4x::facet_wrap2(vars(tag), nrow = 2, labeller = labeller(tag = tag_labels)) +
  ggtitle("UTR Widths", subtitle = "By Gencode Tag") +
  labs(x = "UTR", y = "Width [bp]") +
  custom_gg_theme +
  theme(
    legend.box.just = "left",
    legend.spacing.y = unit(2, "lines"),
    legend.position = "right",
    plot.subtitle = element_text(margin = margin(b = 5, unit = "pt"))
  )
```


```{r table-tag}
utr5_tag <- lapply(valid_tags, function(iTag) unclass(summary(utr_width %>% filter(tag == iTag) %>% pull(utr5_width)))) %>% 
  dplyr::bind_rows() %>% dplyr::mutate(tag = valid_tags)
utr3_tag <- lapply(valid_tags, function(iTag) unclass(summary(utr_width %>% filter(tag == iTag) %>% pull(utr3_width)))) %>% 
  dplyr::bind_rows() %>% dplyr::mutate(tag = valid_tags)

tag_total <- tibble::enframe(table(utr_width$tag)[valid_tags], name = "tag", value = "total")
by_tag_table <- dplyr::bind_rows(
  utr5_tag %>% dplyr::mutate(UTR = "5"),
  utr3_tag %>% dplyr::mutate(UTR = "3")
) %>% 
  dplyr::mutate(across(where(is.numeric), ~round(., 0))) %>% 
  dplyr::filter(!((UTR == "3" & tag == "cds_end_NF") | (UTR == "5" & tag == "cds_start_NF"))) %>% 
  dplyr::left_join(tag_total, by = "tag") %>% 
  dplyr::mutate(UTR = factor(UTR, levels = c("5", "3"))) %>% 
  dplyr::mutate(IQR = `3rd Qu.` - `1st Qu.`) %>% 
  dplyr::select(UTR, tag,
                Median, IQR, `1st Qu.`, `3rd Qu.`, `NA's`, total)
  
by_tag_table %>% 
  printPrettyDf(style = "md") %>% 
  kableExtra::pack_rows(index = table(by_tag_table$UTR))
```


## Split by TSL and Gencode Tag

When the 

```{r plot-tag-tsl, fig.width=12, fig.height=7}
utr_width %>% 
  dplyr::filter(TSL %in% valid_TSL) %>% 
  dplyr::filter(tag %in% valid_tags) %>% 
  dplyr::mutate(TSL = as.factor(TSL)) %>% 
  dplyr::mutate(tag = factor(tag, levels = valid_tags)) %>% 
  tidyr::pivot_longer(cols = c(utr5_width, utr3_width), names_to = "utr", values_to = "utr_width") %>% 
  ggplot(aes(x = utr, y = utr_width, fill = TSL)) +
  geom_boxplot() +
  geom_hline(aes(yintercept = 500, color = "500 bp"), linetype = 5, linewidth = 0.75) +
  scale_color_manual(name = "Txcutr cut-off", values = "black", guide = guide_legend(order = 1)) +
  ggsci::scale_fill_jco(name = "TSL", guide = guide_legend(order = 2)) +
  scale_y_log10() +
  scale_x_discrete(labels = c("utr3_width" = "3'", "utr5_width" = "5'")) +
  ggh4x::facet_wrap2(vars(tag), nrow = 2, labeller = labeller(tag = tag_labels)) +
  ggtitle("UTR Widths", subtitle = "By Gencode Tag and TSL") +
  labs(x = "UTR", y = "Width [bp]") +
  custom_gg_theme +
  theme(
    legend.box.just = "left",
    legend.spacing.y = unit(2, "lines"),
    legend.position = "right",
    plot.subtitle = element_text(margin = margin(b = 5, unit = "pt"))
  )
  
```

# Studies by Transcript {.tabset .break2}

> Section in development

Get the transcript length by exons and by coordinates:

```{r}
tx_width_coord <- transcripts(txdb_ref) %>% 
  tibble::as_tibble() %>%
  dplyr::select(tx_name, coord_width = width)

exons_by_tx <- unlist(exonsBy(txdb_ref, use.names = T))
exons_by_tx$tx_name = names(exons_by_tx)

tx_width_exons <- exons_by_tx %>%
  tibble::as_tibble() %>% 
  dplyr::group_by(tx_name) %>% 
  dplyr::summarise(exon_width = sum(width))

tx_width <- dplyr::left_join(
  tx_width_coord,
  tx_width_exons
)

# Add transcript information
utr_width_mod <- utr_width %>% dplyr::left_join(tx_width, by = c("tx_id" = "tx_name"))
```

```{r, fig.width=9, fig.height=5.5}
plot_hex <- function(utr_width_mod, x_log = T, y_var = "utr3_width", x_var = "exon_width"){
  utr_func <- utr_width_mod %>% 
    dplyr::mutate(y_var = !!sym(y_var),
                  x_var = !!sym(x_var)) %>% 
    dplyr::filter(x_var >= 100) 
  
  if(x_log) utr_func <- utr_func %>% dplyr::mutate(x_var = log10(x_var))
  
  first_hex <- utr_func %>% 
    ggplot(aes(x = x_var, y = y_var)) +
    geom_hex(bins = 250)
    
  xrange <- diff(ggplot_build(first_hex)$layout$panel_params[[1]]$x.range)
  yrange <- diff(ggplot_build(first_hex)$layout$panel_params[[1]]$y.range)
  ratio <- xrange/yrange
  xbins = 250
  xwidth = xrange/xbins
  ywidth = xwidth/ratio
  
  main_hex <- utr_func %>%
    ggplot(aes(x = x_var, y = y_var)) +
    geom_hex(bindwidth = c(xwidth, ywidth)) +
    geom_hline(aes(yintercept = 500, color = "500 bp"), linetype = 5, linewidth = 0.75) +
    viridis::scale_fill_viridis(name = "Number of UTRs in bin") +
    scale_color_manual(name = "Txcutr cut-off", values = "red", guide = guide_legend(order = 1)) +
    coord_fixed(ratio = ratio) +
    labs(x = paste0("Tx. ", x_var), y = y_var) +
    custom_gg_theme +
    theme(
      legend.box.just = "left",
      legend.spacing.y = unit(2, "lines"),
      legend.position = "right",
    )
  
  return(main_hex)
}
```

## 3' UTR width {-}

```{r, fig.width=9, fig.height=5.5}
plot_hex(utr_width_mod, x_log = T, y_var = "utr3_width", x_var = "coord_width")
plot_hex(utr_width_mod, x_log = T, y_var = "utr3_width", x_var = "exon_width")
```

## 5' UTR width {-}

```{r, fig.width=9, fig.height=5.5}
plot_hex(utr_width_mod, x_log = T, y_var = "utr5_width", x_var = "coord_width")
plot_hex(utr_width_mod, x_log = T, y_var = "utr5_width", x_var = "exon_width")
```

# Studies by gene

> Section in development

```{r, fig.width=9, fig.height=5.5}
library(AnnotationDbi)
library(org.Hs.eg.db)

gene_coord_width = genes(txdb_ref) %>%
  tibble::as_tibble() %>% 
  dplyr::select(gene_id, coord_width = width)

exons_by_gene <- unlist(exonsBy(txdb_ref, by = "gene"))
mcols(exons_by_gene)["gene_id"] <- names(exons_by_gene)
gene_exon_width <- exons_by_gene %>% 
  tibble::as_tibble() %>% 
  dplyr::group_by(gene_id) %>% 
  dplyr::summarise(exon_width = sum(width))

# AnnotationDbi::mapIds(org.Hs.eg.db, keys = gene_ids, column = "GENE_LENGTH", keytype = "ENSEMBL", multiVals = "first")

utr_width %>%
  dplyr::left_join(gene_exon_width, by = "gene_id") %>% 
  dplyr::left_join(gene_coord_width, by = "gene_id") %>% 
  ggplot(aes(x = exon_width, y = utr3_width)) +
  geom_abline(intercept = 0, slope = 1, color = "red") +
  geom_point() +
  custom_gg_theme
```


```{r bestagons, fig.width=9, fig.height=5.5}
test <- utr_width %>%
  dplyr::left_join(gene_exon_width, by = "gene_id") %>% 
  dplyr::left_join(gene_coord_width, by = "gene_id") %>% 
  dplyr::filter(exon_width > 100) %>% 
  ggplot(aes(x = log10(exon_width), y = utr3_width)) +
  geom_hex(bins = 100) +
  geom_smooth(method = "lm", color = "red") + 
  custom_gg_theme +
  theme(
    legend.position = "right"
  )

xrange <- diff(ggplot_build(test)$layout$panel_params[[1]]$x.range)
yrange <- diff(ggplot_build(test)$layout$panel_params[[1]]$y.range)
ratio <- xrange/yrange
xbins = 100
xwidth = xrange/xbins
ywidth = xwidth/ratio

utr_width %>%
  dplyr::left_join(gene_exon_width, by = "gene_id") %>% 
  dplyr::left_join(gene_coord_width, by = "gene_id") %>% 
  # dplyr::filter(exon_width > 100) %>% 
  ggplot(aes(x = log10(exon_width), y = utr3_width)) +
  geom_hex(bindwidth = c(xwidth, ywidth)) +
  geom_smooth(method = "lm", color = "red") + 
  coord_fixed(ratio = ratio) +
  custom_gg_theme +
  theme(
    legend.position = "right"
  )
```

```{r, fig.width=9, fig.height=5.5}
utr_width %>%
  dplyr::left_join(gene_exon_width, by = "gene_id") %>% 
  dplyr::left_join(gene_coord_width, by = "gene_id") %>% 
  ggplot(aes(x = exon_width, y = utr3_width)) +
  geom_point(alpha = 0.01) +
  # geom_density2d(color = "blue") +
  custom_gg_theme
  theme(
    legend.position = "right"
  )

df <- utr_width %>%
  dplyr::left_join(gene_exon_width, by = "gene_id") %>% 
  dplyr::left_join(gene_coord_width, by = "gene_id") 
df$gene_bin <- cut(df$exon_width, breaks = quantile(df$exon_width, probs = seq(0, 1, 0.1)), include.lowest = TRUE)
ggplot(df, aes(x = gene_bin, y = utr3_width)) +
  geom_boxplot() +
  labs(x = "Gene Length (Binned)", y = "UTR3 Length") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(df, aes(x = gene_bin, y = utr3_width)) +
  geom_violin() +
  labs(x = "Gene Length Bins", y = "UTR3 Length") +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r}
utr_width %>% 
  dplyr::group_by(gene_id) %>% 
  dplyr::filter(n() > 5) %>% 
  dplyr::summarise(median_3utr = median(utr3_width, na.rm = T)) %>% 
  ggplot(aes(x = gene_id, y = median_3utr)) +
  geom_point() +
  custom_gg_theme
```
