---
title: ENDome Differential Usage
subtitle: satuRn results across pseudobulk presets and satuRn profiles
author:
- name: "Guillermo Rocamora Pérez"
  affiliation: UCL / Cambridge - Version 1.0
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  bookdown::html_document2:
    figure_caption: yes
    code_folding: hide
    theme: paper
    highlight: pygments
    df_print: paged
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
always_allow_html: true
editor_options:
  chunk_output_type: inline
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
shhh <- suppressPackageStartupMessages

shhh({
  library(fs)
  library(conflicted)
  library(tidyverse)
  library(SummarizedExperiment)
})

options(readr.show_progress = FALSE)
options(readr.show_col_types = FALSE)

conflicted::conflict_prefer_all("dplyr", quiet = TRUE)
conflicted::conflict_prefer_all("tidyr", quiet = TRUE)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      out.width = "100%", fig.align = "center", dpi = 150,
                      fig.height = 5.5, fig.width = 9.5)
options(knitr.graphics.error = FALSE)
```

```{r setup-theme, echo = F, results = "asis"}
## Custom sciRmdTheme with increased width.
## Download from devtools::install_github("https://github.com/guillermo1996/grpSciRmdTheme")
if(require(grpSciRmdTheme)){
  grpSciRmdTheme::set.theme(
    theme = "default",
    color = NULL,
    header.sticky = FALSE,
    list.group.icon = "arrow",
    font.family = "Arial",
    font.color = "black",
    header.color = "darkblue",
    figure.perc = 100
  )
}

## A bug in recent versions of KableExtra requires to load the HTML classes in
## some circumstances.
kableExtra::html_dependency_lightable()

cat('
<style type="text/css">
table { caption-side: bottom }
</style>')

theme_endome <- function(base_size = 11, legend_pos = "top"){
  bs <- base_size
  theme_gray(base_size = bs) +
    theme(
      plot.title = element_text(size = bs * 1.27, face = "bold", hjust = 0, vjust = 1, margin = margin(b = bs/2)),
      plot.subtitle = element_text(size = bs*0.82),
      panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5),
      panel.background = element_rect(fill = "#FBFBFB", color = NA),
      axis.text.x = ggplot2::element_text(color = "black", size = bs*0.82, angle = 0, hjust = 0.5),
      axis.text.y = ggplot2::element_text(color = "black", size = bs*0.82),
      axis.title.x = ggplot2::element_text(face = "bold", size = bs, margin = margin(t = 5)),
      axis.title.y = ggplot2::element_text(face = "bold", size = bs, margin = margin(r = 10)),
      panel.grid.minor = element_blank(),
      panel.grid.major.y = element_line(color = "#444444", linewidth = 0.05, linetype = 2, inherit.blank = TRUE),
      panel.grid.major.x = element_line(color = "#888888", linewidth = 0.05, linetype = 1, inherit.blank = TRUE),
      legend.title = element_text(size = bs * 1.09),
      legend.text = element_text(size = bs * 0.82),
      legend.position  = legend_pos,
      legend.key = element_rect(colour = "black", linewidth = 0.01),
      strip.text.x = element_text(color = "black", face = "bold", size = bs*0.82),
      strip.background = element_rect(color = "black", linewidth = 0.5, linetype = "solid"),
      strip.text.y = element_text(color = "black", face = "bold", size = bs*0.82),
      plot.margin = margin(0.5, 0.5, 0.2, 0.5, "cm")
    )
}
```

```{r snakemake-preamble, echo = FALSE}
# Define Snakemake interactive parameters ----
if (!exists("snakemake")) {
  library(methods)
  Snakemake <- setClass(
    "Snakemake",
    slots = c(
      input = 'list',
      output = 'list',
      params = 'list',
      wildcards = 'list',
      log = 'list',
      threads = 'numeric',
      scriptdir = 'character')
  )
  ## dataset.group.merge_method folder created by the `test_data` rule (its
  ## name is the prefix wildcard) and the ENDome to test. Every satuRn run linked
  ## for that ENDome (test.<endome>.<pseudobulk>.<profile>[.satuRn].condition.*) is picked
  ## up from the folder, with the pseudobulk of its preset
  ## (test.<endome>.<pseudobulk>.pseudobulk.rds) for the usage plots. Absolute
  ## paths, as knitr runs from the Rmd folder.
  repo_dir <- "~/RytenLab-Research/40-ENDome_generation"
  test_dir <- file.path(repo_dir, "data/test_data/Ebbert.control.st_ref")
  test_orf_filter <- "ref_pc"
  test_width <- "500"
  test_txEnd <- "3p"
  test_endome <- paste0(test_orf_filter, ".w", test_width, ".", test_txEnd)
  test_pattern <- paste0("^test\\.", gsub(".", "\\.", test_endome, fixed = TRUE), "\\.(.+)\\.%s$")

  snakemake <- Snakemake(
    input = list(
      results = list.files(test_dir, pattern = sprintf(test_pattern, "condition\\.tsv\\.gz"), full.names = TRUE),
      diagnostics = list.files(test_dir, pattern = sprintf(test_pattern, "condition\\.diagnostics\\.tsv"), full.names = TRUE),
      pseudobulk = list.files(test_dir, pattern = sprintf(test_pattern, "pseudobulk\\.rds"), full.names = TRUE),
      sample_covariates = file.path(repo_dir, "data/Hardy_snRNAseq/annotations/Hardy_final_covariates_weighted_and_combined_batches.csv")
    ),
    output = list(
      html = file.path(test_dir, paste0("interactive.", test_endome, ".satuRn_report.html"))
    ),
    params = list(
      ## The whole pseudobulk_settings and satuRn_settings blocks of the config:
      ## only used to show the settings of the runs found in the file names
      pseudobulk_settings = list(
        default = list(effective_txs_mode = "fixed", min_effective_txs = 300, cell_type_level = "annotation_level_2"),
        dynamic = list(effective_txs_mode = "mad", mad_below_median = 3, cell_type_level = "annotation_level_2")
      ),
      satuRn_settings = list(
        default = list(contrasts = c("PDD-Control", "PD-Control", "PDD-PD"), min_cells = 5, p_value = "raw", alpha = 0.05,
                       covariates_sample_col = "sample", group_col = "group")
      ),
      genes_of_interest = c("APP", "SNCA"),
      top_genes = 6
    ),
    wildcards = list(
      endome_name = paste0(basename(test_dir), ".", test_endome)
    ),
    threads = 4,
    scriptdir = file.path(repo_dir, "workflow/scripts")
  )
}
```

# Introduction

This report summarizes the differential usage results of `07c-satuRn_DTU.R` for a single ENDome, and compares the different **runs** that were produced for it. A run is a combination of a pseudobulk preset (`pseudobulk_settings`, how nuclei are selected and grouped in `07a`/`07b`) and a satuRn profile (`satuRn_settings`, how the fits are set up and tested in `07c`). Each run is read from the name of its result files (`<endome>.<pseudobulk>.<profile>.condition.tsv.gz` and `.condition.diagnostics.tsv`), so every table and figure has one tab per run, and new runs appear without changing the report.

The report covers five questions:

1. **Which fits ran.** How many region x cell type fits reached satuRn, with how many samples, genes and bins, and why the others were skipped.
2. **How well calibrated are the tests.** The empirical null estimated by satuRn and the p-value distributions, which decide how much of the signal survives.
3. **What is differentially used.** Significant genes per cell type and contrast, their effect sizes, and in which direction the usage moves along the gene (towards the upstream or the downstream ENDome end).
4. **The genes of interest and the top hits,** with the bin usage of every sample.
5. **How the runs agree.** The overlap and the rank concordance of every run with the others.

A gene is significant when its stage-wise adjusted p-value from stageR (`stageR_gene`) is below the `alpha` of its satuRn profile, and a bin when its confirmation-stage p-value (`stageR_transcript`) is.

# Hidden Section {.unlisted .unnumbered .hidden-heading .tabset .break4}

## File paths {-}

```{r script-paths}
# Snakemake inputs
smk_inputs <- snakemake@input
smk_outputs <- snakemake@output
params <- snakemake@params
wc <- snakemake@wildcards

pseudobulk_settings <- params$pseudobulk_settings
satuRn_settings <- params$satuRn_settings
genes_of_interest <- unlist(params$genes_of_interest)
top_genes <- params$top_genes %||% 6

#' Run of an input file, read from its name
#'
#' Files are named <endome>.<run>.<suffix>, with <endome> ending in
#' .w<width>.<txEnd> (e.g. Ebbert.control.st_ref.ref_pc.w500.3p.default.satuRn.condition.tsv.gz
#' or test.ref_pc.w500.3p.dynamic.pseudobulk.rds).
runFromPath <- function(path, suffix) {
  if (length(path) == 0) return(stats::setNames(character(0), character(0)))
  run <- sub(sprintf("^.*\\.w[^.]+\\.\\dp\\.(.+)\\.%s$", gsub(".", "\\.", suffix, fixed = TRUE)), "\\1", basename(path))
  if (any(run == basename(path))) stop("Cannot read the run from: ", paste(path[run == basename(path)], collapse = ", "), call. = FALSE)
  stats::setNames(path.expand(path), run)
}

## Runs: <pseudobulk preset>.<satuRn profile>, matched by name between results and diagnostics
## (07c names its outputs <endome>.<pseudobulk>.<profile>.satuRn.condition.*: the
## ".satuRn" tag is not part of the run)
results_paths <- runFromPath(unlist(smk_inputs$results), "condition.tsv.gz")
diagnostics_paths <- runFromPath(unlist(smk_inputs$diagnostics), "condition.diagnostics.tsv")
names(results_paths) <- sub("\\.satuRn$", "", names(results_paths))
names(diagnostics_paths) <- sub("\\.satuRn$", "", names(diagnostics_paths))
pseudobulk_paths <- runFromPath(unlist(smk_inputs$pseudobulk), "pseudobulk.rds")
covariates_path <- path.expand(smk_inputs$sample_covariates %||% smk_inputs$covariates)

runs <- intersect(names(results_paths), names(diagnostics_paths))
unpaired <- setdiff(union(names(results_paths), names(diagnostics_paths)), runs)
if (length(unpaired) > 0) warning("Runs without both results and diagnostics (left out): ", paste(unpaired, collapse = ", "))
if (length(runs) == 0) stop("No run has both results and diagnostics", call. = FALSE)

run_info <- tibble::tibble(
  run = runs,
  pseudobulk = sub("\\..*$", "", runs),
  profile = dplyr::if_else(grepl("\\.", runs), sub("^[^.]+\\.", "", runs), "-"),
  label = paste0(pseudobulk, " / ", profile)
)
run_labels <- stats::setNames(run_info$label, run_info$run)

### Truncation site of the ENDome: which end defines the bins
tx_end <- stringr::str_extract(wc$endome_name, "(3p|5p)$")

### Helper library
lib_dir <- path.expand(file.path(snakemake@scriptdir, "../lib"))
if(!dir.exists(lib_dir)) lib_dir <- path.expand(file.path(snakemake@scriptdir, "lib"))
source(file.path(lib_dir, "hf-notebooks.R"))
```

```{r script-paths-table, echo = FALSE}
dplyr::bind_rows(
  tibble::tibble(Run = run_labels[runs], Input = "Results", Path = results_paths[runs]),
  tibble::tibble(Run = run_labels[runs], Input = "Diagnostics", Path = diagnostics_paths[runs]),
  tibble::tibble(Run = names(pseudobulk_paths), Input = "Pseudobulk", Path = unname(pseudobulk_paths)),
  tibble::tibble(Run = "-", Input = "Sample covariates", Path = covariates_path)
) %>%
  renderTable(style = "md", caption = "Inputs of this report.")
```

## Helper functions {-}

```{r helper-functions}
fmtInt <- function(x) format(round(x), big.mark = ",", trim = TRUE, scientific = FALSE)
fmtPct <- function(x, accuracy = 0.1) scales::percent(x, accuracy = accuracy)
fmtP <- function(x) dplyr::if_else(is.na(x), "-", formatC(x, format = "g", digits = 3))

#' Write one tab per run
#'
#' Must be called from a chunk with `results = "asis"`, right below a header
#' with the `.tabset` class. `fun(run)` returns a ggplot, a kable or NULL.
tabsetByRun <- function(fun, level = 3) {
  for (r in runs) {
    cat("\n\n", strrep("#", level), " ", run_labels[[r]], " {-}\n\n", sep = "")
    out <- fun(r)
    if (inherits(out, "ggplot")) {
      print(out)
    } else if (inherits(out, "knitr_kable") || inherits(out, "kableExtra")) {
      cat(as.character(out))
    } else if (!is.null(out)) {
      print(out)
    }
    cat("\n\n")
  }
}

#' Settings of a run: the pseudobulk preset and the satuRn profile, if in the config
runSetting <- function(r, key, default = NULL) {
  profile <- run_info$profile[run_info$run == r]
  satuRn_settings[[profile]][[key]] %||% default
}

#' Index of the minimum, NA when every value is NA
whichMinSafe <- function(x) if (all(is.na(x))) NA_integer_ else which.min(x)

#' -log10 with a floor, so p = 0 can be drawn
nlog10 <- function(p) -log10(pmax(p, 1e-300))
```

## Data loading {-}

Results, diagnostics, gene-level summaries and gene symbols for every run.

```{r load-data}
## Results (one row per run x region x cell type x contrast x bin) and diagnostics (one row per fit x contrast)
results <- lapply(runs, function(r) {
  x <- readr::read_tsv(results_paths[[r]])
  if ("regular_FDR" %in% colnames(x)) x <- dplyr::rename(x, FDR = regular_FDR)
  x
}) %>% stats::setNames(runs)
diagnostics <- lapply(runs, function(r) readr::read_tsv(diagnostics_paths[[r]])) %>% stats::setNames(runs)
alpha <- vapply(runs, function(r) as.numeric(runSetting(r, "alpha", 0.05)), numeric(1))

## Gene symbols from org.Hs.eg.db when available; the gene id otherwise (e.g. novel MSTRG genes)
all_gene_ids <- unique(unlist(lapply(results, function(x) x$gene_id)))
gene_symbol <- stats::setNames(all_gene_ids, all_gene_ids)
if (requireNamespace("org.Hs.eg.db", quietly = TRUE)) {
  ens <- sub("\\..*$", "", all_gene_ids)
  sym <- suppressMessages(AnnotationDbi::mapIds(org.Hs.eg.db::org.Hs.eg.db, keys = unique(ens[grepl("^ENSG", ens)]),
                                                keytype = "ENSEMBL", column = "SYMBOL", multiVals = "first"))
  gene_symbol[] <- dplyr::coalesce(unname(sym[ens]), all_gene_ids)
}

## Gene-level summary of every test: smallest bin p-value, stageR gene p-value, bins
genes <- lapply(runs, function(r) {
  results[[r]] %>%
    dplyr::group_by(region, celltype, contrast, gene_id) %>%
    dplyr::summarise(
      n_bins = dplyr::n(),
      min_p = suppressWarnings(min(p_used, na.rm = TRUE)),
      gene_qvalue = dplyr::first(gene_qvalue),
      stageR_gene = dplyr::first(stageR_gene),
      n_sig_bins = sum(stageR_transcript < alpha[[r]], na.rm = TRUE),
      best_bin = isoform_id[whichMinSafe(p_used)][1],
      best_estimate = estimates[whichMinSafe(p_used)][1],
      .groups = "drop"
    ) %>%
    dplyr::mutate(min_p = dplyr::if_else(is.finite(min_p), min_p, NA_real_),
                  significant = !is.na(stageR_gene) & stageR_gene < alpha[[r]],
                  symbol = unname(gene_symbol[gene_id]))
}) %>% stats::setNames(runs)

## Shared cell type order (most bins tested first) and contrasts
celltype_levels <- dplyr::bind_rows(results) %>% dplyr::count(celltype, sort = TRUE) %>% dplyr::pull(celltype)
celltype_levels <- union(celltype_levels, unique(unlist(lapply(diagnostics, function(d) d$celltype))))
contrast_levels <- unique(unlist(lapply(diagnostics, function(d) d$contrast)))

## Sample groups and pseudobulks, for the usage plots
samples <- readr::read_csv(covariates_path)
samples <- samples %>%
  dplyr::rename(sample_id = dplyr::all_of(runSetting(runs[1], "covariates_sample_col", "sample")),
                group = dplyr::all_of(runSetting(runs[1], "group_col", "group"))) %>%
  dplyr::select(sample_id, group)
pseudobulks <- lapply(pseudobulk_paths, readRDS)
```

# Runs {.tabset .tabset-pills}

The settings of each run: its pseudobulk preset and its satuRn profile, as given in the config. A preset or profile that is not in the config is reported as such (its results are still shown).

```{r runs-table, results = "asis", echo = FALSE}
tabsetByRun(level = 2, function(r) {
  info <- run_info[run_info$run == r, ]
  settingsTbl <- function(s, block) {
    if (is.null(s)) return(tibble::tibble(Block = block, Setting = "-", Value = "not in the config"))
    tibble::tibble(Block = block, Setting = names(s),
                   Value = vapply(s, function(v) paste(unlist(v), collapse = ", "), character(1)))
  }
  dplyr::bind_rows(
    settingsTbl(pseudobulk_settings[[info$pseudobulk]], paste0("pseudobulk: ", info$pseudobulk)),
    settingsTbl(satuRn_settings[[info$profile]], paste0("satuRn: ", info$profile))
  ) %>%
    renderTable(style = "md", caption = sprintf("Settings of the `%s` run.", info$label))
})
```

# Fits

`07c` runs one fit per region x cell type, with every group of the contrasts, and tests the contrasts from it. A fit is skipped when a group has too few samples with enough nuclei, or when too few genes survive the `DRIMSeq::dmFilter` step.

## Summary {.tabset .tabset-pills}

```{r fits-summary, results = "asis", echo = FALSE}
tabsetByRun(function(r) {
  d <- diagnostics[[r]]
  ok <- d %>% dplyr::filter(status == "ok")
  sig <- genes[[r]] %>% dplyr::filter(significant)
  tibble::tibble(
    Metric = c("Tests (fit x contrast)", "Tests that ran", "Tests skipped",
               "Region x cell types that ran", "Genes tested (sum over tests)", "Bins tested (sum over tests)",
               "Tests using empirical p-values", "Tests using raw p-values",
               "Significant genes (sum over tests)", "Distinct significant genes", "Significant bins (sum over tests)"),
    Value = c(fmtInt(nrow(d)), fmtInt(nrow(ok)), fmtInt(sum(d$status != "ok")),
              fmtInt(dplyr::n_distinct(paste(ok$region, ok$celltype))),
              fmtInt(sum(ok$n_genes_tested, na.rm = TRUE)), fmtInt(sum(ok$n_bins_tested, na.rm = TRUE)),
              fmtInt(sum(ok$p_type == "empirical", na.rm = TRUE)), fmtInt(sum(ok$p_type == "raw", na.rm = TRUE)),
              fmtInt(nrow(sig)), fmtInt(dplyr::n_distinct(sig$gene_id)),
              fmtInt(sum(results[[r]]$stageR_transcript < alpha[[r]], na.rm = TRUE)))
  ) %>%
    renderTable(style = "md", caption = sprintf("Fits of the `%s` run (alpha = %s).", run_labels[[r]], alpha[[r]]))
})
```

## Skipped tests {.tabset .tabset-pills}

```{r fits-skipped, results = "asis", echo = FALSE}
tabsetByRun(function(r) {
  d <- diagnostics[[r]] %>% dplyr::filter(status != "ok")
  if (nrow(d) == 0) { cat("No test was skipped.\n\n"); return(NULL) }
  d %>%
    dplyr::mutate(reason = sub(" \\(.*$", "", status)) %>%
    dplyr::count(region, reason, name = "tests") %>%
    dplyr::rename(Region = region, Reason = reason, Tests = tests) %>%
    renderTable(style = "md", caption = sprintf("Skipped tests of the `%s` run, by reason.", run_labels[[r]]))
})
```

## Fits per region and cell type {.tabset .tabset-pills}

One row per test that ran: the samples per group, the genes and bins tested, which p-values were used, the empirical null estimated by satuRn (`delta`, `sigma`), the fraction of bins with a raw p-value below 0.05, and the significant genes and bins.

```{r fits-table, results = "asis", echo = FALSE}
tabsetByRun(function(r) {
  diagnostics[[r]] %>%
    dplyr::filter(status == "ok") %>%
    dplyr::mutate(celltype = factor(celltype, levels = celltype_levels),
                  frac_raw_p05 = fmtPct(frac_raw_p05), null_delta = round(null_delta, 3), null_sigma = round(null_sigma, 3)) %>%
    dplyr::arrange(region, celltype, contrast) %>%
    dplyr::select(Region = region, `Cell type` = celltype, Contrast = contrast, `Samples per group` = group_sizes,
                  Genes = n_genes_tested, Bins = n_bins_tested, `P-values` = p_type, Delta = null_delta,
                  Sigma = null_sigma, `Raw p < 0.05` = frac_raw_p05, `Sig. genes` = n_sig_genes, `Sig. bins` = n_sig_bins) %>%
    renderTable(style = "md", caption = sprintf("Tests of the `%s` run.", run_labels[[r]]))
})
```

# Calibration

satuRn computes, for every contrast with at least 500 bins, an *empirical* null: a normal distribution with mean `delta` and standard deviation `sigma` fitted to the bulk of the test statistics. With a well-calibrated test it is close to N(0, 1). A `sigma` above 1 means the raw statistics are more spread than the theory expects. The empirical p-values remove that excess, which protects against miscalibration but also removes a real, broad shift. Which p-values each test used is in the `P-values` column above (`p_value` in the satuRn profile).

```{r calibration-sigma, results = "asis", echo = FALSE, fig.height = 7}
cat("\n\n## Empirical null spread {.tabset .tabset-pills}\n\n")
cat("`sigma` of every test with an empirical null. The dashed line is a perfectly calibrated test (sigma = 1).\n\n")
tabsetByRun(function(r) {
  d <- diagnostics[[r]] %>% dplyr::filter(status == "ok", !is.na(null_sigma))
  if (nrow(d) == 0) { cat("No test has an empirical null (all have fewer than 500 bins).\n\n"); return(NULL) }
  d %>%
    dplyr::mutate(celltype = factor(celltype, levels = celltype_levels)) %>%
    ggplot(aes(x = null_sigma, y = forcats::fct_rev(celltype), colour = contrast, shape = region)) +
    geom_vline(xintercept = 1, linetype = 2) +
    geom_point(size = 2.5, position = position_dodge(width = 0.5)) +
    labs(x = "Empirical null sigma", y = NULL, colour = "Contrast", shape = "Region",
         title = "Spread of the empirical null", subtitle = sprintf("Run: %s", run_labels[[r]])) +
    theme_endome()
})
```

```{r calibration-pvalues, results = "asis", echo = FALSE, fig.height = 6}
cat("\n\n## P-value distributions {.tabset .tabset-pills}\n\n")
cat("Raw and empirical bin p-values of every test, by contrast. Under no signal both are flat; an excess near 0 in the raw p-values that disappears in the empirical ones is what the empirical correction removes.\n\n")
tabsetByRun(function(r) {
  results[[r]] %>%
    dplyr::select(contrast, Raw = pval, Empirical = empirical_pval) %>%
    tidyr::pivot_longer(c(Raw, Empirical), names_to = "type", values_to = "p") %>%
    dplyr::filter(!is.na(p)) %>%
    dplyr::mutate(type = factor(type, levels = c("Raw", "Empirical"))) %>%
    ggplot(aes(x = p)) +
    geom_histogram(breaks = seq(0, 1, 0.025), fill = "#2166ac", colour = "white", linewidth = 0.1) +
    facet_grid(type ~ contrast, scales = "free_y") +
    labs(x = "Bin p-value", y = "Bins", title = "P-value distributions", subtitle = sprintf("Run: %s", run_labels[[r]])) +
    theme_endome()
})
```

# Differential usage

## Significant genes per test {.tabset .tabset-pills}

Number of significant genes per region, cell type and contrast. Grey tiles are tests that were skipped.

```{r dtu-counts, results = "asis", echo = FALSE, fig.height = 7.5}
tabsetByRun(function(r) {
  diagnostics[[r]] %>%
    dplyr::mutate(celltype = factor(celltype, levels = celltype_levels),
                  n_sig_genes = dplyr::if_else(status == "ok", n_sig_genes, NA_real_)) %>%
    ggplot(aes(x = contrast, y = forcats::fct_rev(celltype), fill = n_sig_genes)) +
    geom_tile(colour = "white") +
    geom_text(aes(label = dplyr::if_else(is.na(n_sig_genes), "", as.character(n_sig_genes))), size = 3) +
    scale_fill_gradient(low = "#F7FBFF", high = "#08519C", na.value = "grey85", trans = "pseudo_log") +
    facet_wrap(~ region) +
    labs(x = NULL, y = NULL, fill = "Significant\ngenes", title = "Significant genes per test",
         subtitle = sprintf("Run: %s | stageR gene p-value < %s", run_labels[[r]], alpha[[r]])) +
    theme_endome() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
})
```

```{r dtu-swarm, results = "asis", echo = FALSE, fig.height = 9}
cat("\n\n## Gene significance per cell type {.tabset .tabset-pills}\n\n")
cat("Every tested gene, by cell type: the stage-wise adjusted gene p-value (stageR) on a -log10 scale, one panel per contrast and region. Coloured genes are significant; the labels mark the most significant gene of each cell type and the genes of interest.\n\n")
tabsetByRun(function(r) {
  g <- genes[[r]] %>%
    dplyr::filter(!is.na(stageR_gene)) %>%
    dplyr::mutate(celltype = factor(celltype, levels = celltype_levels), y = nlog10(stageR_gene))
  labels <- dplyr::bind_rows(
    g %>% dplyr::filter(significant) %>% dplyr::group_by(region, contrast, celltype) %>% dplyr::slice_max(y, n = 1, with_ties = FALSE) %>% dplyr::ungroup(),
    g %>% dplyr::filter(symbol %in% genes_of_interest, significant)
  ) %>% dplyr::distinct()
  ggplot(g, aes(x = celltype, y = y)) +
    geom_hline(yintercept = -log10(alpha[[r]]), linetype = 2, colour = "red") +
    geom_jitter(data = dplyr::filter(g, !significant), colour = "grey70", width = 0.25, height = 0, size = 0.5) +
    geom_jitter(data = dplyr::filter(g, significant), colour = "#b2182b", width = 0.25, height = 0, size = 1.2) +
    ggrepel::geom_text_repel(data = labels, aes(label = symbol), size = 2.6, max.overlaps = 30, min.segment.length = 0) +
    facet_grid(contrast ~ region) +
    labs(x = NULL, y = "-log10 stageR gene p-value", title = "Gene significance per cell type",
         subtitle = sprintf("Run: %s | dashed line: alpha = %s", run_labels[[r]], alpha[[r]])) +
    theme_endome() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
})
```

## Significant genes {.tabset .tabset-pills}

The significant genes of each run, most significant first (at most 200 rows). `Best bin` is the bin with the smallest p-value and `Log-odds` its satuRn estimate: the log-odds ratio of its usage between the two groups of the contrast (positive = more used in the first group).

```{r dtu-table, results = "asis", echo = FALSE}
tabsetByRun(function(r) {
  sig <- genes[[r]] %>% dplyr::filter(significant)
  if (nrow(sig) == 0) { cat("No significant gene in this run.\n\n"); return(NULL) }
  sig %>%
    dplyr::arrange(stageR_gene, min_p) %>%
    dplyr::slice_head(n = 200) %>%
    dplyr::transmute(Region = region, `Cell type` = celltype, Contrast = contrast, Gene = symbol, `Gene id` = gene_id,
                     Bins = n_bins, `Sig. bins` = n_sig_bins, `stageR gene p` = fmtP(stageR_gene),
                     `Best bin` = best_bin, `Min. p` = fmtP(min_p), `Log-odds` = round(best_estimate, 2)) %>%
    renderTable(style = "md", caption = sprintf("Significant genes of the `%s` run.", run_labels[[r]]))
})
```

```{r dtu-volcano, results = "asis", echo = FALSE, fig.height = 8}
cat("\n\n## Effect sizes {.tabset .tabset-pills}\n\n")
cat("Bin log-odds ratio against its p-value for the four cell types with most bins tested, one row per contrast. Red bins are significant after the stage-wise correction.\n\n")
tabsetByRun(function(r) {
  top_ct <- results[[r]] %>% dplyr::count(celltype, sort = TRUE) %>% dplyr::slice_head(n = 4) %>% dplyr::pull(celltype)
  results[[r]] %>%
    dplyr::filter(celltype %in% top_ct, !is.na(p_used)) %>%
    dplyr::mutate(celltype = factor(celltype, levels = top_ct),
                  significant = !is.na(stageR_transcript) & stageR_transcript < alpha[[r]]) %>%
    dplyr::arrange(significant) %>%
    ggplot(aes(x = estimates, y = nlog10(p_used), colour = significant)) +
    geom_point(size = 0.5, alpha = 0.6) +
    scale_colour_manual(values = c(`FALSE` = "grey70", `TRUE` = "#b2182b"), labels = c(`FALSE` = "Not significant", `TRUE` = "Significant")) +
    facet_grid(contrast ~ celltype + region) +
    guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1))) +
    labs(x = "Log-odds ratio (satuRn estimate)", y = "-log10 p-value", colour = NULL,
         title = "Bin effect sizes", subtitle = sprintf("Run: %s", run_labels[[r]])) +
    theme_endome()
})
```

## Direction of the usage shift {.tabset .tabset-pills}

```{r dtu-direction-text, results = "asis", echo = FALSE}
end_label <- if (identical(tx_end, "5p")) "5' end (TSS)" else "3' end (polyadenylation site)"
cat(sprintf(
  "Where the significant bins sit within their gene, and whether they are more or less used in the first group of the contrast. Bins are ranked by their %s in the direction of transcription: the *most upstream* bin of a gene is %s, the *most downstream* %s, and the rest are *internal*.\n\n",
  end_label,
  if (identical(tx_end, "5p")) "the most distal start" else "the shortest 3' UTR (most proximal site)",
  if (identical(tx_end, "5p")) "the most proximal start" else "the longest 3' UTR (most distal site)"))
```

```{r dtu-direction, results = "asis", echo = FALSE, fig.height = 6}
tabsetByRun(function(r) {
  x <- results[[r]]
  if (!all(c("is_most_upstream_end", "is_most_downstream_end") %in% colnames(x))) {
    cat("The results have no bin positions (the pseudobulk was built without the ENDome GTF).\n\n"); return(NULL)
  }
  x <- x %>%
    dplyr::filter(!is.na(stageR_transcript), stageR_transcript < alpha[[r]]) %>%
    dplyr::mutate(position = dplyr::case_when(is_most_upstream_end ~ "Most upstream",
                                              is_most_downstream_end ~ "Most downstream",
                                              TRUE ~ "Internal"),
                  position = factor(position, levels = c("Most upstream", "Internal", "Most downstream")),
                  direction = dplyr::if_else(estimates > 0, "More used in group 1", "Less used in group 1"))
  if (nrow(x) == 0) { cat("No significant bin in this run.\n\n"); return(NULL) }
  x %>%
    ggplot(aes(x = position, fill = direction)) +
    geom_bar(position = "dodge", colour = "black", linewidth = 0.2) +
    scale_fill_manual(values = c(`More used in group 1` = "#b2182b", `Less used in group 1` = "#2166ac")) +
    facet_wrap(~ contrast) +
    labs(x = "Position of the bin's end within the gene", y = "Significant bins", fill = NULL,
         title = "Direction of the usage shift", subtitle = sprintf("Run: %s | group 1 = first group of the contrast", run_labels[[r]])) +
    theme_endome()
})
```

# Bin usage

Proportion of each gene's counts that falls in each of its bins, per pseudobulk sample (donor x region) and group, in the cell type and region where the gene is most significant. Only samples with at least the profile's `min_cells` nuclei are shown, as in the fits. Bins are labelled by the rank of their ENDome end (`E1` = most upstream) when the pseudobulk carries the bin positions.

```{r usage-helpers}
#' Per-sample bin usage of a set of (region, cell type, gene) tests, from the run's pseudobulk
usageData <- function(r, tests) {
  pb <- pseudobulks[[run_info$pseudobulk[run_info$run == r]]]
  if (is.null(pb) || nrow(tests) == 0) return(NULL)
  min_cells <- as.numeric(runSetting(r, "min_cells", 5))
  cell_counts <- if (requireNamespace("dreamlet", quietly = TRUE)) dreamlet::cellCounts(pb) else NULL
  rd <- as.data.frame(rowData(pb))
  purrr::pmap_dfr(tests, function(region, celltype, gene_id, symbol, ...) {
    bins <- rownames(pb)[rd$gene_id == gene_id]
    if (length(bins) < 2 || !celltype %in% assayNames(pb)) return(NULL)
    keep <- stringr::str_extract(colnames(pb), "[^_]+$") == region & colnames(pb) %in% samples$sample_id
    if (!is.null(cell_counts)) keep <- keep & cell_counts[colnames(pb), celltype] >= min_cells
    X <- as.matrix(assay(pb, celltype)[bins, keep, drop = FALSE])
    X <- X[, colSums(X) > 0, drop = FALSE]
    if (ncol(X) == 0) return(NULL)
    bin_label <- if ("end_rank" %in% colnames(rd)) paste0("E", rd[bins, "end_rank"]) else bins
    tibble::tibble(
      panel = paste0(symbol, " - ", celltype, " - ", region),
      bin = rep(bin_label, times = ncol(X)),
      sample_id = rep(colnames(X), each = length(bins)),
      usage = as.vector(sweep(X, 2, colSums(X), "/"))
    ) %>% dplyr::left_join(samples, by = "sample_id")
  })
}

usagePlot <- function(u, title, subtitle) {
  if (is.null(u) || nrow(u) == 0) return(NULL)
  ggplot(u, aes(x = bin, y = usage, fill = group)) +
    geom_boxplot(outlier.shape = NA, linewidth = 0.3, alpha = 0.7) +
    geom_point(position = position_jitterdodge(jitter.width = 0.15), size = 0.6, alpha = 0.6) +
    scale_y_continuous(labels = scales::percent) +
    facet_wrap(~ panel, scales = "free_x") +
    labs(x = "Bin", y = "Share of the gene's counts", fill = "Group", title = title, subtitle = subtitle) +
    theme_endome()
}
```

```{r usage-goi, results = "asis", echo = FALSE, fig.height = 5.5}
cat("\n\n## Genes of interest {.tabset .tabset-pills}\n\n")
cat(sprintf("Genes of interest: %s. The table lists every test of these genes; the figure shows their usage in the test where each is most significant.\n\n",
            paste(genes_of_interest, collapse = ", ")))
tabsetByRun(function(r) {
  g <- genes[[r]] %>% dplyr::filter(symbol %in% genes_of_interest)
  if (nrow(g) == 0) { cat("None of the genes of interest was tested in this run.\n\n"); return(NULL) }
  ranks <- genes[[r]] %>%
    dplyr::group_by(region, celltype, contrast) %>%
    dplyr::mutate(rank = rank(min_p, ties.method = "min", na.last = "keep"), n_genes = dplyr::n()) %>%
    dplyr::ungroup() %>% dplyr::select(region, celltype, contrast, gene_id, rank, n_genes)
  tbl <- g %>%
    dplyr::left_join(ranks, by = c("region", "celltype", "contrast", "gene_id")) %>%
    dplyr::arrange(symbol, stageR_gene, min_p) %>%
    dplyr::transmute(Gene = symbol, Region = region, `Cell type` = celltype, Contrast = contrast,
                     `stageR gene p` = fmtP(stageR_gene), `Min. p` = fmtP(min_p),
                     Rank = paste0(rank, " / ", n_genes), `Sig. bins` = n_sig_bins, Significant = significant) %>%
    renderTable(style = "md", caption = sprintf("Genes of interest in the `%s` run.", run_labels[[r]]))
  cat(as.character(tbl))
  best <- g %>% dplyr::group_by(symbol) %>% dplyr::slice_min(min_p, n = 1, with_ties = FALSE) %>% dplyr::ungroup()
  usagePlot(usageData(r, best %>% dplyr::select(region, celltype, gene_id, symbol)),
            "Bin usage of the genes of interest", sprintf("Run: %s | test with the smallest p-value of each gene", run_labels[[r]]))
})
```

```{r usage-top, results = "asis", echo = FALSE, fig.height = 7}
cat(sprintf("\n\n## Top %s genes {.tabset .tabset-pills}\n\n", top_genes))
cat("The most significant tests of each run (by stageR gene p-value, then smallest bin p-value), one gene per test.\n\n")
tabsetByRun(function(r) {
  top <- genes[[r]] %>%
    dplyr::filter(!is.na(min_p)) %>%
    dplyr::arrange(stageR_gene, min_p) %>%
    dplyr::distinct(gene_id, .keep_all = TRUE) %>%
    dplyr::slice_head(n = top_genes)
  usagePlot(usageData(r, top %>% dplyr::select(region, celltype, gene_id, symbol)),
            "Bin usage of the top genes", sprintf("Run: %s", run_labels[[r]]))
})
```

# Comparison between runs

How each run agrees with the others, over the tests (region x cell type x contrast) and genes that both ran. The rank concordance uses the smallest bin p-value of each gene, which is available whatever the p-values used for the significance calls.

## Overlap {.tabset .tabset-pills}

```{r compare-table, results = "asis", echo = FALSE}
keys <- c("region", "celltype", "contrast", "gene_id")
tabsetByRun(function(r) {
  others <- setdiff(runs, r)
  if (length(others) == 0) { cat("Only one run: nothing to compare.\n\n"); return(NULL) }
  purrr::map_dfr(others, function(o) {
    j <- dplyr::inner_join(genes[[r]], genes[[o]], by = keys, suffix = c("_r", "_o"))
    sig_r <- sum(genes[[r]]$significant); sig_o <- sum(genes[[o]]$significant)
    both <- sum(j$significant_r & j$significant_o)
    tibble::tibble(
      `Compared with` = run_labels[[o]],
      `Shared tests` = dplyr::n_distinct(paste(j$region, j$celltype, j$contrast)),
      `Shared genes tested` = nrow(j),
      `Sig. here` = sig_r, `Sig. there` = sig_o, `Sig. in both` = both,
      Jaccard = round(both / max(1, sig_r + sig_o - both), 3),
      `Spearman (min. p)` = round(suppressWarnings(stats::cor(j$min_p_r, j$min_p_o, method = "spearman", use = "complete.obs")), 3)
    )
  }) %>%
    renderTable(style = "md", caption = sprintf("Agreement of the `%s` run with the other runs (significant gene x test pairs).", run_labels[[r]]))
})
```

```{r compare-scatter, results = "asis", echo = FALSE, fig.height = 6}
cat("\n\n## Gene significance against the other runs {.tabset .tabset-pills}\n\n")
cat("Smallest bin p-value of every gene tested in both runs (-log10), this run on the x axis, coloured by where the gene is significant.\n\n")
tabsetByRun(function(r) {
  others <- setdiff(runs, r)
  if (length(others) == 0) { cat("Only one run: nothing to compare.\n\n"); return(NULL) }
  purrr::map_dfr(others, function(o) {
    dplyr::inner_join(genes[[r]], genes[[o]], by = keys, suffix = c("_r", "_o")) %>%
      dplyr::mutate(other = run_labels[[o]])
  }) %>%
    dplyr::filter(!is.na(min_p_r), !is.na(min_p_o)) %>%
    dplyr::mutate(status = factor(dplyr::case_when(significant_r & significant_o ~ "Both",
                                                   significant_r ~ "This run only",
                                                   significant_o ~ "Other run only",
                                                   TRUE ~ "Neither"),
                                  levels = c("Neither", "Other run only", "This run only", "Both"))) %>%
    dplyr::arrange(status) %>%
    ggplot(aes(x = nlog10(min_p_r), y = nlog10(min_p_o), colour = status)) +
    geom_abline(slope = 1, intercept = 0, linetype = 2) +
    geom_point(size = 0.6, alpha = 0.6) +
    scale_colour_manual(values = c(Neither = "grey75", `Other run only` = "#ef8a62", `This run only` = "#67a9cf", Both = "#b2182b")) +
    facet_wrap(~ other) +
    guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1))) +
    labs(x = "-log10 min. bin p-value (this run)", y = "-log10 min. bin p-value (other run)", colour = "Significant in",
         title = "Gene significance across runs", subtitle = sprintf("Run: %s", run_labels[[r]])) +
    theme_endome()
})
```

# Session info {.unlisted .unnumbered}

```{r session-info, echo = FALSE}
sessionInfo()
```
